---
title: "Introduction to Alztool: Standardizing Clinical Data"
date: "r Sys.Date()"
author: "Merhawi Mussa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to alztool}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview
The Alztool package provides standardized, reproducible functions for transforming raw clinical variables (like BMI and MMSE scores) into categorized features commonly used in Alzheimer's Disease and health informatics research. This vignette demonstrates the package's core functionality using a simulated dataset.

## Installation
The package can be installed directly from GitHub using the devtools package:

 Install devtools if necessary
 install.packages("devtools") 

devtools::install_github("mussamerhawi/Alztool")


### 1. Load the Package and Simulate Data
First, we load the package and create a small sample dataset that mimics the structure of your full Alzheimer's dataset.
```{r}
library(Alztool)
library(dplyr)
library(knitr)

# Simulate a small dataset for demonstration
sim_data <- data.frame(
  patient_id = 101:105,
  bmi = c(17.9, 23.5, 28.1, 32.0, 39.9),
  mmse = c(29, 21, 25, 18, NA),
  # Other columns to match your real dataset structure
  age = c(70, 75, 68, 80, 72) 
)

# Display the starting data
kable(sim_data)
```

### 2. Standardize Cognitive Status (standardize_mmse)
We use the standardize_mmse function to classify subjects as "Impaired," "Not Impaired," or "Missing" based on a cutoff score of 24.
```{r}
data_mmse_classified <- sim_data %>%
  standardize_mmse()

# Display the results
kable(data_mmse_classified)

```

### 3. Classify Body Mass Index (classify_bmi)
Next, we apply the classify_bmi function to categorize the raw BMI scores into standard WHO categories (Underweight, Normal, Overweight, Obese Class 1/2/3).
```{r}
final_classified_data <- data_mmse_classified %>%
  classify_bmi()

# Select only the key columns for easy viewing
final_classified_data <- final_classified_data %>%
  select(patient_id, bmi, bmi_category, mmse, cognitive_status)

# Display the final results, showing the success of the classification
kable(final_classified_data)
```

## Summary
The Alztool package successfully transforms raw numeric inputs into categorical variables (bmi_category and cognitive_status), providing a reliable and reproducible first step for clinical data analysis.
